Flatband ferromagnetism and dispersion relations of spin-1 excitations in topological Hubbard models

### 2018-06-14 paper/paper.tex
Version 1.1

\section{Introduction}
\par Electronic bands with nonzero topological indices reside on the center of a substantial amount of topological phenomena in condensed matter physics \cite{HK_RMP2010,QZ_RMP2011}. It was proposed in a pioneering work by Haldane \cite{H_PRL1988} that a spinless fermionic model on a honeycomb lattice exhibits integer quantum Hall effect \cite{KDP_PRL1980} without an external magnetic field. This model, serving as the first example of Chern insulator, breaks the time reversal symmetry with a complex next nearest neighbor hopping and is characterized by a nonzero Chern number \cite{TKNN_PRL1982}. Later, the concept was generalized to time reversal symmetric systems with spin-orbital coupling (SOC), such as the monolayer graphene \cite{KM_PRL2005a,KM_PRL2005b} and HgTe/CdTe quantum wells \cite{BHZ_S2006,Ketc_S2007}. The SOC there generates complex hopping terms similar to that proposed by Haldane but with opposite chiralities for electrons with up spins and down spins, resulting in the quantum spin Hall insulator characterized by a $Z_2$ index.

\par The lattice models with nontrivial band topology share much similarity with the two-dimensional electron gas (2DEG) under a strong magnetic field with Landau levels, e.g. the existence of topologically protected gapless edge states \cite{H_PRL1988,L_PRB1981,H_PRB1982,KM_PRL2005a,KM_PRL2005b,BHZ_S2006,Ketc_S2007}. Thus more novel phases other than the Chern insulator or $Z_2$ insulator are expected when Coulomb interactions are taken into account, as is similar to the fractional quantum Hall effect \cite{TSG_PRL1982,L_PRL1983} in the 2DEG with Landau levels. However, different from Landau levels, energy bands in lattice models usually have noneligible dispersions, which weakens the effect of Coulomb interactions. Therefore, in recent years, much effort has been devoted to the design and search of tight-binding models that host nearly-flat electron bands with nontrivial topology \cite{TMW_PRL2011,WR_PRB2011,SGKD_PRL2011,WGGS_PRL2011,NSCM_PRL2011,TB_PRB2012,YGSD_PRB2012}. Analogous exotic phases, such as the fractional Chern insulator and fractional topological insulator were numerically verified to emerge in such nearly-flat topological bands when strong Coulomb interactions are turned on \cite{NSCM_PRL2011,SGSS_NC2011,RB_PRX2011,NSRCM_PRB2011,LBFL_PRL2012}.

@@

\par We consider a generalized $\pi$-flux Hubbard model on the square lattice, whose Hamiltonian can be written as $H=H_0+H_U$, where $H_0$ is the generalization with electron spin of the original spinless model proposed in Ref. \cite{NSCM_PRL2011},
\begin{equation}
H_0=\sum_{\langle ij\rangle,\sigma}(t_1^{ij,\sigma}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.})
    +\sum_{\langle\langle ij\rangle\rangle,\sigma}(t_2^{ij}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.}),
\end{equation}
and $H_U$ is the Hubbard interaction
\begin{equation}
H_U=U\sum_{i}n_{i\uparrow}n_{i\downarrow}.
\end{equation}
Here, $c_{i\sigma}^{\dagger}(c_{i\sigma})$ creates (annihilates) a spin $\sigma$ electron at site $i$, $n_{i\sigma}$ is the spin $\sigma$ electron number operator at site $i$, $\langle ij\rangle$ denotes the nearest-neighbor (NN) bonds and $\langle\langle ij\rangle\rangle$ denotes the next-nearest-neighbor (NNN) bonds. As is shown in Fig. \ref{lattice}(a), the spin-dependent NN hopping amplitude $t_1^{ij,\sigma}$ and the spin-independent NNN hopping amplitude $t_2^{ij}$ are given by
\begin{equation}
t_1^{ij,\sigma}=t_1\exp\left[i(-1)^{\delta^1_{ij}}\alpha_\sigma\pi/4\right],
\end{equation}
\begin{equation}
t_2^{ij}=t_2(-1)^{\delta^2_{ij}}.
\end{equation}
Here, $\delta^1_{ij}=+1$ if the NN electron hopping is along the direction of the solid black arrow and $\delta^1_{ij}=-1$ if along the reversed direction. $\delta^2_{ij}=+1$ if the NNN electron hopping is along the dashed green lines and $\delta^2_{ij}=-1$ if along the dashed purple lines. The spin-dependent phase $\alpha_\sigma$ breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=\alpha_\downarrow=+1$, whereas it preserves the time-reversal symmetry but breaks the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=+1$ and $\alpha_\downarrow=-1$.

@@

\par When there is a gap between the $d$ band and the $f$ band, these bands can be shown to be topologically non-trivial by calculating their Chern numbers \cite{TKNN_PRL1982} (for the time-reversal-symmetry breaking case) or $Z_2$ indices \cite{KM_PRL2005b,SWSH_PRL2006} (for the time-reversal-symmetry preserving case). The Chern number for a single spin component of the $d$ band or the $f$ band can be expressed in terms of the coefficients $D_{i,\mathbf{k}}$ \cite{HK_RMP2010,QZ_RMP2011},

@@

\section{Numerical results}\label{nr}
\begin{figure}
\includegraphics[width=\columnwidth]{phase}
\caption{(Color online) Phase diagram of the quarter-filled (a) Chern Hubbard model and (b) $Z_2$ Hubbard model. The colormap represents the ratio $U/W$ of the lower electron band, where $U$ is the Hubbard interaction strength and $W$ is the lower electron bandwidth. Red stars in (a) and (b) mark the parameters used in Fig. \ref{fmcispectrum} and Fig. \ref{tispectrum}(a), respectively. Blue stars in (a) and (b) mark the parameters used in Fig. \ref{pbcispectrum}. and Fig. \ref{tispectrum}(b), respectively.}
\label{phase}
\end{figure}

@@

\subsection{Chern Hubbard model}\label{CHM}
\begin{figure}
\includegraphics[width=\columnwidth]{fmcispectrum}
\caption{(Color online) (a) Spin-1 excitation spectra of the Chern Hubbard model in the FM phase. Green solid lines denote the spin waves and the grey region denotes the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Blue solid lines denote the upper and lower boundaries of the Stoner continuum determined by the split band picture shown in (c). Black arrows mark the local minima of the Stoner continuum. (b) Spectral weights for the lowest four eigen levels of the spin-1 excitation spectra with the center-of-mass momentum $\mathbf{q}=(0,0)$. (c) Illustration of the splitting of the lower electron bands for up spins (blue solid line) and down spins (blue dashed line) in the presence of ferromagnetism of the Chern Hubbard model. Inset shows the positions of the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Green arrows denote the corresponding scattering channels of the dips of the Stoner continuum marked in (a). (d) Differences of the spectral weights subtracted by those in the flatband limit for the lowest four eigen levels of the spin-1 excitation spectra with the center-of-mass momentum $\mathbf{q}=(\pi,0)$. The parameters are $t_1=1.0$, $t_2=0.49$ and $U=2.3$.}
\label{fmcispectrum}
\end{figure}

@@

\par To explore the physics related to the nonflatness of the lower electron band, the corresponding spin-1 excitation spectra in the flatband limit are also shown in Fig. \ref{fmcispectrum}(a) as the green dashed lines. The flatband-limit spectra consist of two parts as well, i.e. the low-lying spin waves and the high-energy Stoner continuum. However, in this case, the Stoner continuum is just a pile of featureless flat lines at $U/2$. Comparing the spectra of the original model with this, as indicated by the black arrows in Fig. \ref{fmcispectrum}(a), we can conclude that the nonflatness of the electron band introduces dips of the lower boundary of the Stoner continuum. Furthermore, around the valleys of these dips, the energies of the collective modes are apparently renormalized downward, leading to the occurrence of roton-like local minima of the optical branch.

\par The above physics can be illustrated by a simple band-splitting picture as shown in Fig. \ref{fmcispectrum}(c). Roughly speaking, in the presence of ferromagnetism, the spin-down electron band and spin-up electron band are split by an energy gap proportional to $U\langle m \rangle$, where $U$ is the Hubbard interaction strength and $\langle m \rangle$ is the average magnetic moment per site. For a quarter-filled electron model with two inequivalent sublattices, $\langle m \rangle=1/2$. Thus the energy splitting is $U/2$. Then an individual spin-1 excitation, or a mode in the Stoner continuum, can be viewed as the excitation of an electron from the lower fully-filled spin-up band to the upper empty spin-down band. This argument can be verified by our numerical results. In fact, the sharp locality on the $\mathbf{k}_i$ space of the spectral weights of a mode in the Stoner continuum [as has be seen in Fig. \ref{fmcispectrum}($\text{b}_3$) and Fig. \ref{fmcispectrum}($\text{b}_4)$] implies that this kind of mode is mainly determined by the corresponding diagonal term in Eq. \ref{PHP}. By numerics, we also find that the term $\frac{U}{N} \sum_{a=1,2} \sum_{\mathbf{p}\neq\mathbf{k}_i} \left| \mu_{a,\mathbf{p}\uparrow} \right|^2 \left| \mu_{a,\mathbf{k}_{i}-\mathbf{q}\downarrow} \right|^2 \sim \frac{U}{2}$. Therefore, the excitation energies of the Stoner continuum can be approximated by $\varepsilon_{d\downarrow} (\mathbf{k}_i-\mathbf{q}) + \frac{U}{2} -\varepsilon_{d\uparrow} (\mathbf{k}_i)$, which is consistent with the above picture. In Fig. \ref{fmcispectrum}(a), the lower and upper boundaries of the Stoner continuum approximated by this argument are plotted as the blue solid lines, which fits extremely well with the true boundaries of the shaded area, thus verifies the validity of this simple argument. Now we can find out the dominant scattering channels that contribute most to the dips of the Stoner continuum and read off their center-of-mass momentums easily. In the inset of Fig. \ref{fmcispectrum}(c), we mark the local maxima of the spin-up band with red solid circles and the local minima of the spin-down band with purple ones. The scattering channels corresponding to the valley bottoms marked by the black arrows in Fig. \ref{fmcispectrum}(a) are also indicated by green arrows. Obviously, a spin-1 excitation of an electron from one of the spin-up maxima to one of the spin-down minima appears as a valley bottom of the lower boundary of the Stoner continuum and the momentum difference of these two electron state provides the center-of-mass momentum of that spin-1 excitation.

\par Within this picture, the collective modes of the spectra can be viewed as bound states in the gap of the Stoner continuum. Then the physics of the downward renormalization of the energies of the collective modes induced by the dips of the Stoner continuum is quite similar to that observed in the two-dimensional antiferromagents, where the suppression of the energies of the spin waves are explained by their couplings to the nearby deconfined spinon continuum \cite{ZFSMC_PRL2006,TS_PRL2013,DMCNTPEMIR_NP2015,YWDYL_arXiv2018}. To understand the renormalization of the collective modes observed here, in Fig. \ref{fmcispectrum}(d), we plot the spectral weights of the four lowest eigen levels subtracted by those in the flatband limit with the center-of-mass momentum $q=(\pi,0)$ ($X_1$ point). The differences concentrate around the scattering channels labeled by $\mathbf{k}_i=(\pm\pi,0)$ and $\mathbf{k}_i=(0,\pm\pi)$. However, the results for the collective modes [Fig. \ref{fmcispectrum}($\text{d}_1$) and Fig. \ref{fmcispectrum}($\text{d}_2$)] are positive while those for the individual modes at the valley bottom of the Stoner continuum [Fig. \ref{fmcispectrum}($\text{d}_3$) and Fig. \ref{fmcispectrum}($\text{d}_4$)] are negative. Therefore a noticeable amount of spectral weights transfer from the latter to the former, which indicates a strong coupling between the spin waves and the nearby Stoner continuum. This coupling renormalizes the energies of the spin waves and leads to the occurrence of a roton-like local minimum for the optical branch.

@@

\par In this subsection, we focus on the quarter-filled $Z_2$ Hubbard model. Its spin-1 excitation spectra along a high symmetry path in the first Brillouin zone with the parameters marked by the red star and blue star in Fig. \ref{phase}(b) are plotted in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b), respectively. As is similar to Sec. \ref{CHM}, the low-lying green solid lines represent the spin waves (collective modes) and the high-energy shaded areas represent the Stoner continuum (individual modes). A distinction from the Chern Hubbard model is that the collective modes here are gapped because the $Z_2$ Hubbard model explicitly breaks the spin $SU(2)$ rotation symmetry and no spontaneous continuous symmetry breaking occurs in the ferromagnetic phase. The corresponding spectra in the flatband limit are shown as the green dashed lines in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b) as well for comparison. Similarly, the nonflatness of the electron band introduces dips of the Stoner continuum, and the energies of the spin waves around the valley bottoms of the Stoner continuum are renormalized downward, which results in roton-like modes on both branches of the spin waves. The dominant scattering channels leading to the dips of the Stoner continuum and the renormalization of the spin waves can also be determined by the previously proposed band-splitting picture, which is illustrated in Fig. \ref{tispectrum}(c) and Fig. \ref{tispectrum}(d). Their insets shows the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Here only the scattering channels that contribute to the valley bottoms marked by the black arrows in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b) are stressed because their induced roton-like modes lead to the destabilization of the ferromagnetic ground state [See Fig. \ref{tispectrum}(b)]. We want to remark that although the center-of-mass momentums of the forementioned valley bottoms are the same, the scattering channels are different because the bottom of the spin-down band changes from the $\Gamma$ and $M_{1,2,3,4}$ points to the $X_{1,2}$ and $Y_{1,2}$ points when $t_2$ changes from $0.65$ to $0.741$ [See Fig. \ref{tispectrum}(c) and Fig. \ref{tispectrum}(d)]. Last but not least, the ferromagnetic ground state is also quite robust against the spin flips until the bottom of the spin-down band almost overlaps with the spin-up band on energy, which means the phase boundary of the FM/NFM is also quite close to $U/W=2$.

>>>>

\section{Introduction}
\par Electronic bands with nonzero topological indices reside on the center of a substantial amount of topological phenomena in condensed matter physics \cite{HK_RMP2010,QZ_RMP2011}. It was proposed in a pioneering work by Haldane \cite{H_PRL1988} that a spinless fermionic model on a honeycomb lattice exhibits integer quantum Hall effect \cite{KDP_PRL1980} without an external magnetic field. This model, serving as the first example of Chern insulator, breaks the time reversal symmetry with a complex next-nearest-neighbor hopping and is characterized by a nonzero Chern number \cite{TKNN_PRL1982}. Later, the concept was generalized to time reversal symmetric systems with spin-orbital coupling (SOC), such as the monolayer graphene \cite{KM_PRL2005a,KM_PRL2005b} and HgTe/CdTe quantum wells \cite{BHZ_S2006,Ketc_S2007}. The SOC there generates complex hopping terms similar to that proposed by Haldane but with opposite chiralities for electrons with up spins and down spins, resulting in the quantum spin Hall insulator characterized by a $Z_2$ index.

\par The lattice models with nontrivial band topology share much similarity with the two-dimensional electron gas (2DEG) under a strong magnetic field with Landau levels, e.g. the existence of topologically protected gapless edge states \cite{H_PRL1988,L_PRB1981,H_PRB1982,KM_PRL2005a,KM_PRL2005b,BHZ_S2006,Ketc_S2007,YXL_PRL2011,LYGL_PRB2016,LGLW_NJP2017}. Thus more novel phases other than the Chern insulator or $Z_2$ insulator are expected when Coulomb interactions are taken into account, as is similar to the fractional quantum Hall effect \cite{TSG_PRL1982,L_PRL1983} in the 2DEG with Landau levels. However, different from Landau levels, energy bands in lattice models usually have noneligible dispersions, which weakens the effect of Coulomb interactions. Therefore, in recent years, much effort has been devoted to the design and search of tight-binding models that host nearly-flat electron bands with nontrivial topology \cite{TMW_PRL2011,WR_PRB2011,SGKD_PRL2011,WGGS_PRL2011,NSCM_PRL2011,TB_PRB2012,YGSD_PRB2012}. Analogous exotic phases, such as the fractional Chern insulator and fractional topological insulator were numerically verified to emerge in such nearly-flat topological bands when strong Coulomb interactions are turned on \cite{NSCM_PRL2011,SGSS_NC2011,RB_PRX2011,NSRCM_PRB2011,LBFL_PRL2012}.

@@

\par We consider a generalized $\pi$-flux Hubbard model on the square lattice, whose Hamiltonian can be written as $H=H_0+H_U$, where $H_0$ is the generalization with electron spin of the original spinless model proposed in Ref. \cite{NSCM_PRL2011},
\begin{equation}
H_0=\sum_{\langle ij\rangle,\sigma}(t_1^{ij,\sigma}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.})
    +\sum_{\langle\langle ij\rangle\rangle,\sigma}(t_2^{ij}c_{i\sigma}^\dagger c_{j\sigma}+\text{H.c.}),
\end{equation}
and $H_U$ is the Hubbard interaction
\begin{equation}
H_U=U\sum_{i}n_{i\uparrow}n_{i\downarrow}.
\end{equation}
Here, $c_{i\sigma}^{\dagger}(c_{i\sigma})$ creates (annihilates) a spin $\sigma$ electron at site $i$, $n_{i\sigma}$ is the spin $\sigma$ electron number operator at site $i$, $\langle ij\rangle$ denotes the nearest-neighbor (NN) bonds and $\langle\langle ij\rangle\rangle$ denotes the next-nearest-neighbor (NNN) bonds. As is shown in Fig. \ref{lattice}(a), the spin-dependent NN hopping amplitude $t_1^{ij,\sigma}$ and the spin-independent NNN hopping amplitude $t_2^{ij}$ are given by
\begin{equation}
t_1^{ij,\sigma}=t_1\exp\left[i\delta^1_{ij}\alpha_\sigma\pi/4\right],
\end{equation}
\begin{equation}
t_2^{ij}=t_2\delta^2_{ij}.
\end{equation}
Here, $\delta^1_{ij}=+1$ if the NN electron hopping is along the direction of the solid black arrow and $\delta^1_{ij}=-1$ if along the reversed direction. $\delta^2_{ij}=+1$ if the NNN electron hopping is along the dashed green lines and $\delta^2_{ij}=-1$ if along the dashed purple lines. The spin-dependent phase $\alpha_\sigma$ breaks the time-reversal symmetry but preserves the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=\alpha_\downarrow=+1$, whereas it preserves the time-reversal symmetry but breaks the spin $SU(2)$ rotation symmetry if $\alpha_\uparrow=+1$ and $\alpha_\downarrow=-1$.

@@

\par When there is a gap between the $d$ band and the $f$ band, these bands can be shown to be topologically non-trivial by calculating their Chern numbers \cite{TKNN_PRL1982} (for the time-reversal-symmetry-breaking case) or $Z_2$ indices \cite{KM_PRL2005b,SWSH_PRL2006} (for the time-reversal-symmetry-preserving case). The Chern number for a single spin component of the $d$ band or the $f$ band can be expressed in terms of the coefficients $D_{i,\mathbf{k}}$ \cite{HK_RMP2010,QZ_RMP2011},

@@

\section{Numerical results}\label{nr}
\begin{figure}
\includegraphics[width=\columnwidth]{phase}
\caption{(Color online) Phase diagrams of the quarter-filled (a) Chern Hubbard model and (b) $Z_2$ Hubbard model. The colormap represents the ratio $U/W$ of the lower electron band, where $U$ is the Hubbard interaction strength and $W$ is the lower electron bandwidth. Red stars in (a) and (b) mark the parameters used in Fig. \ref{fmcispectrum} and Fig. \ref{tispectrum}(a), respectively. Blue stars in (a) and (b) mark the parameters used in Fig. \ref{pbcispectrum}. and Fig. \ref{tispectrum}(b), respectively.}
\label{phase}
\end{figure}

@@

\subsection{Chern Hubbard model}\label{CHM}
\begin{figure}
\includegraphics[width=\columnwidth]{fmcispectrum}
\caption{(Color online) (a) Spin-1 excitation spectra of the Chern Hubbard model in the FM phase. Green solid lines denote the spin waves and the grey region denotes the Stoner continuum. Green dashed lines represent the corresponding spectra in the flatband limit. Blue solid lines denote the upper and lower boundaries of the Stoner continuum determined by the band-splitting picture shown in (c). Black arrows mark the local minima of the Stoner continuum. (b) Spectral weights for the lowest four eigen levels of the spin-1 excitation spectra with the center-of-mass momentum $\mathbf{q}=(0,0)$. (c) Illustration of the splitting of the lower electron bands for up spins (blue solid line) and down spins (blue dashed line) in the presence of ferromagnetism of the Chern Hubbard model. Inset shows the positions of the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Green arrows denote the corresponding scattering channels of the valley bottoms of the Stoner continuum marked in (a). (d) Differences of the spectral weights subtracted by those in the flatband limit for the lowest four eigen levels of the spin-1 excitation spectra with the center-of-mass momentum $\mathbf{q}=(\pi,0)$. The parameters are $t_1=1.0$, $t_2=0.49$ and $U=2.3$.}
\label{fmcispectrum}
\end{figure}

@@

\par To explore the physics related to the nonflatness of the lower electron band, the corresponding spin-1 excitation spectra in the flatband limit are also shown in Fig. \ref{fmcispectrum}(a) as the green dashed lines. The flatband-limit spectra consist of two parts as well, i.e. the low-lying spin waves and the high-energy Stoner continuum. However, in this case, the Stoner continuum is just a pile of featureless flat lines at $U/2$. Comparing the spectra of the original model with this, as indicated by the black arrows in Fig. \ref{fmcispectrum}(a), we can conclude that the nonflatness of the electron band introduces dips of the lower boundary of the Stoner continuum. Furthermore, around these dips, the energies of the collective modes are apparently renormalized downward, leading to the occurrence of roton-like local minima on the optical branch.

\par The above physics can be illustrated by a simple band-splitting picture as shown in Fig. \ref{fmcispectrum}(c). Roughly speaking, in the presence of ferromagnetism, the spin-down electron band and spin-up electron band are split in energy proportional to $U\langle m \rangle$, where $U$ is the Hubbard interaction strength and $\langle m \rangle$ is the average magnetic moment per site. For a quarter-filled electron model with two inequivalent sublattices, $\langle m \rangle=1/2$. Thus the energy splitting is $U/2$. Then an individual spin-1 excitation, or a mode in the Stoner continuum, can be viewed as the excitation of an electron from the lower fully-filled spin-up band to the upper empty spin-down band. This argument can be verified by our numerical results. In fact, the sharp locality on the $\mathbf{k}_i$ space of the spectral weights of a mode in the Stoner continuum [as has be seen in Fig. \ref{fmcispectrum}($\text{b}_3$) and Fig. \ref{fmcispectrum}($\text{b}_4)$] implies that this kind of mode is mainly determined by the corresponding diagonal term in Eq. \ref{PHP}. By numerics, we also find that the term $\frac{U}{N} \sum_{a=1,2} \sum_{\mathbf{p}\neq\mathbf{k}_i} \left| \mu_{a,\mathbf{p}\uparrow} \right|^2 \left| \mu_{a,\mathbf{k}_{i}-\mathbf{q}\downarrow} \right|^2 \sim \frac{U}{2}$. Therefore, the excitation energies of the Stoner continuum can be approximated by $\varepsilon_{d\downarrow} (\mathbf{k}_i-\mathbf{q}) + \frac{U}{2} -\varepsilon_{d\uparrow} (\mathbf{k}_i)$, which is consistent with the above picture. In Fig. \ref{fmcispectrum}(a), the lower and upper boundaries of the Stoner continuum approximated by this argument are plotted as the blue solid lines, which fits extremely well with the true boundaries of the shaded area, thus verifies the validity of this simple argument. Now we can find out the dominant scattering channels that contribute most to the valley bottoms of the Stoner continuum and read off their center-of-mass momentums easily. In the inset of Fig. \ref{fmcispectrum}(c), we mark the local maxima of the spin-up band with red solid circles and the local minima of the spin-down band with purple ones. The scattering channels corresponding to the valley bottoms marked by the black arrows in Fig. \ref{fmcispectrum}(a) are also indicated by green arrows. Obviously, a spin-1 excitation of an electron from one of the spin-up maxima to one of the spin-down minima appears as a valley bottom of the lower boundary of the Stoner continuum and the momentum difference of these two electron state provides the center-of-mass momentum of that spin-1 excitation.

\par Within this picture, the collective modes of the spectra can be viewed as bound states in the gap of the Stoner continuum. Then the physics of the downward renormalization of the energies of the collective modes induced by the valley bottoms of the Stoner continuum is quite similar to that observed in the two-dimensional antiferromagents, where the suppression of the energies of the spin waves are explained by their couplings to the nearby deconfined spinon continuum \cite{ZFSMC_PRL2006,TS_PRL2013,DMCNTPEMIR_NP2015,YWDYL_arXiv2018}. To understand the renormalization of the collective modes observed here, in Fig. \ref{fmcispectrum}(d), we plot the spectral weights of the four lowest eigen levels subtracted by those in the flatband limit with the center-of-mass momentum $q=(\pi,0)$ ($X_1$ point). The differences concentrate around the scattering channels labeled by $\mathbf{k}_i=(\pm\pi,0)$ and $\mathbf{k}_i=(0,\pm\pi)$. However, the results for the collective modes [Fig. \ref{fmcispectrum}($\text{d}_1$) and Fig. \ref{fmcispectrum}($\text{d}_2$)] are positive while those for the individual modes at the valley bottom of the Stoner continuum [Fig. \ref{fmcispectrum}($\text{d}_3$) and Fig. \ref{fmcispectrum}($\text{d}_4$)] are negative. Therefore a noticeable amount of spectral weights transfer from the latter to the former, which indicates a strong coupling between the spin waves and the nearby Stoner continuum. This coupling renormalizes the energies of the spin waves and leads to the occurrence of a roton-like local minimum for the optical branch.

@@

\par In this subsection, we focus on the quarter-filled $Z_2$ Hubbard model. Its spin-1 excitation spectra along a high symmetry path in the first Brillouin zone with the parameters marked by the red star and blue star in Fig. \ref{phase}(b) are plotted in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b), respectively. As is similar to Sec. \ref{CHM}, the low-lying green solid lines represent the spin waves (collective modes) and the high-energy shaded areas represent the Stoner continuum (individual modes). A distinction from the Chern Hubbard model is that the collective modes here are gapped because the $Z_2$ Hubbard model explicitly breaks the spin $SU(2)$ rotation symmetry and no spontaneous continuous symmetry breaking occurs in the ferromagnetic phase. The corresponding spectra in the flatband limit are shown as the green dashed lines in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b) as well for comparison. Similarly, the nonflatness of the electron band introduces dips of the Stoner continuum, and the energies of the spin waves around the valley bottoms of the Stoner continuum are renormalized downward, which results in roton-like modes on both branches of the spin waves. The dominant scattering channels leading to the dips of the Stoner continuum and the renormalization of the spin waves can also be determined by the previously proposed band-splitting picture, which is illustrated in Fig. \ref{tispectrum}(c) and Fig. \ref{tispectrum}(d). Their insets shows the corresponding local maxima (red solid circles) of the spin-up bands and local minima (purple solid circles) of the spin-down bands in the first Brillouin zone. Here only the scattering channels that contribute to the valley bottoms marked by the black arrows in Fig. \ref{tispectrum}(a) and Fig. \ref{tispectrum}(b) are stressed because their induced roton-like modes lead to the destabilization of the ferromagnetic ground state [See Fig. \ref{tispectrum}(b)]. We want to remark that although the center-of-mass momentums of the forementioned valley bottoms are the same, the scattering channels are different because the bottom of the spin-down band changes from the $\Gamma$ and $M_{1,2,3,4}$ points to the $X_{1,2}$ and $Y_{1,2}$ points when $t_2$ changes from $0.65$ to $0.741$ [See Fig. \ref{tispectrum}(c) and Fig. \ref{tispectrum}(d)]. Last but not least, the ferromagnetic ground state is also quite robust against the spin flips until the bottom of the spin-down band almost overlaps with the top of the spin-up band on energy, which means the phase boundary of the FM/NFM is also quite close to $U/W=2$.
